\part{What is Ethics? Moral Relativism Vs Moral Objectivism}
\label{ch.modone}
\addtocontents{toc}{\protect\mbox{}\protect\hrulefill\par}
\chapter{Part 1: What is Ethics? What is Relativism?}
Much like the other sciences, Philosophy is broken up in to fields of study which share certain aspects concerning methodology. There are many different fields which one could specialize in but each one has three distinct levels of inquiry. These levels, roughly, in order of abstraction, are 'meta', 'normative/theoretical', and 'applied'. Starting with the most abstract, the 'meta' level concerns questions about the questions in the field. So, for example, metabiology would be an area of biology asking questions such as 'what exactly is a species?', 'where is the line between biology and chemestry?' and others of that sort. Meta-ethics, which is the content for this module and the next, asks questions of the form 'what do right and wrong even mean?', 'what are we doing when we make ethical claims?', 'is any action, regardless of circumstance, absolutely wrong?', and 'concerning actions, what is the difference between 'right and wrong' and 'good and bad'?'. The theoretical (for Ethics, the term is "normative") level concerns the hypotheses and explanations for various things in the field. Generally, these questions/explanations are tested and used in the applied level. Normative Ethics concerns questions like "what general feature(s) make some action right or wrong?", "what does it take to be a good person?", and others like that.  The final, lowest level, is the applied level. This is where they look at real world cases, actually test the hypotheses in the meta and normative levels, to see what happens. For Ethics, this is found in questions like "what should I do in this situation?". The majority of this class concerns Normative Ethics, with some adventures into the other levels. 

Ethics, fundamentally speaking, is a field of study which concerns value, duty, right, wrong, justice, fairness, obligations, and so on.  Within Normative Ethics, there are two 'branches', so to speak, which we will explore in this textbook. The first concerns questions regarding individual actions, so, basically, "what general feature makes an action right or wrong?". This is where we will spend most of our time as it's more practically useful and many of the motivations behind contemporary civil rights movements can be found there. The second branch concerns what makes a good or bad person. Basically, "what general features make a good person?". You will likely see me use this terminology again but the first branch concerns what you should do and the second concerns who you should be. These are the titular questions of this textbook: `Who Should I Be?' and `What Should I Do?'.

\section{Going Meta: The Ethics Family Tree}  

You may have sometimes heard  the term “meta” used and attached it to a field of study, such as “metaphysics”, “metaethics”, “metabiology”, and even “metaphilosophy”. The prefix “meta-” changes the level of questioning in the field. Physics is trying to answer certain questions, metaphysics is trying to answer questions about the questions in physics. If it helps, the prefix “meta-” means “asking questions about the questions in”, and then every additional “meta-” moves the level up a peg. The 'lowest' level of abstraction for a field is its applied version. So, you have physics and then you have applied physics, mathematics and applied mathematics. The next level up concerns abstractions from the findings in the applied field. This is either called 'theoretical' or (in the case of Ethics) 'normative'. And, finally, the most abstract level which most people need to be concerned about is 'meta-'. So, using this way of understanding, I can make the following chart: 

\begin{center}
\begin{tikzpicture}[modal]
		\node[world] (w1) [label=right:What is Ethics?] {Meta-Ethics};
		\node[world] (w2) [label=right:What Makes Actions Right or Wrong?, below=of w1]{Normative Ethics};
		\node[world] (w3) [label=right:What About These Real-World Cases?, below=of w2]{Applied Ethics};
		\path[->][label=right:What do `Right' and `Wrong' even mean?] (w1) edge (w2);
		\path[->] (w2) edge (w3);
	\end{tikzpicture}
\end{center}

Relativism and Objectivism are stances in the meta-level of Ethics. But there are two others, Nihilism and Skepticism, which are on the same level. One can think of the stances in Ethics as falling on a family tree. The stances are related to each other according to how they answer certain questions. For simplicity's sake, I have arranged these according to yes-no questions, but it is possible to get similar relations by going with degrees. The first question is 'are there moral facts?', that is 'are moral statements either true or false?'. Both Relativism and Objectivism answer the first question in the same way, they say 'yes' so they are called 'realist' stances. Nihilism says 'no' and Skepticism answers it by saying 'I don't know'. The Objectivist and the Relativist differ in that the objectivist claims that the morality of something doesn't change according to the culture or the person. Opinions about what is right or wrong may change, but the Objectivist says that what is actually moral/immoral is constant. People were just wrong about what was OK to do. The Relativist says that morality is determined by the culture/person, so when perspectives change, morality itself changes (this will be a problem for the theory later). 

\begin{center}
\begin{tikzpicture}[modal]
		\node[world] (w1) [label=right:] {\tiny{Are There Moral Facts?}};
		\node[world] (w2) [below left=of w1]{Yes};
		\node[world] (w3) [below=of w1]{No};
		\node[world] (w4) [below right=of w1]{No Clue};
		\node[world] (w5) [below=of w2]{\tiny{Moral Realism}};
		\node[world] (w6) [below=of w3]{\tiny{Moral Nihilism}};
		\node[world] (w7) [below=of w4]{\tiny{Moral Skepticism}};
		\node[world] (w8) [below=of w5]{\tiny{Do they change?}};
		\node[world] (w9) [below left=of w8]{No};
		\node[world] (w11) [below=of w9]{\tiny{Moral Objectivism}};
		\path[->] (w9) edge (w11);
		\node[world] (w10) [below right=of w8]{Yes};
		\node[world] (w12) [below=of w10]{\tiny{Moral Relativism}};
		\path[->] (w8) edge (w9);
		\path[->] (w8) edge (w10);
		\path[->] (w9) edge (w11);
		\path[->] (w10) edge (w12);
		\path[->] (w5) edge (w8);

		\path[->] (w1) edge (w2);
		\path[->] (w1) edge (w3);
		\path[->] (w1) edge (w4);
		\path[->] (w2) edge (w5);
		\path[->] (w3) edge (w6);
		\path[->] (w4) edge (w7);

	\end{tikzpicture}
\end{center}


\section{Relativism, in General}

Though we will be spending most of our time handling questions concerning Ethical/Moral Relativism, it is useful to go through exactly what that term 'Relativism' means. This sort of idea is found in several areas of philosophy and the notion (applied differently) is found in physics. This, in its broad sense, is the stance that there is no absolute truth (objective truth), but rather the truth about things varies from person to person or from culture to culture, depending on the version of relativism which is held. Much like any other theory, there are several different 'relativisms' which differ according to their scope as well as the aspect which the facts in that scope are relative to.  This may be that there are no objective truths at all, or that there aren’t any in some particular area. Relativism comes in many different flavors. The first is that everything is relative, this is global relativism. The other is that only certain things are relative, this is limited relativism. From this, we have to ask what are they relative to? These are either to your culture or to you individually. These two divisions into relativism gives us a total of 4 different 'classes' of relativism which one could find themselves in.

First, we have Global Cultural Relativism. This says that all truths, all facts, are determined by what your culture believes. So, for example, if this were true, there would be no fact of the matter about whether the Earth is flat, rather it would be flat for one culture (The flat earthers) and round for another. This same thing would apply to all other facts, such as mathematics. In a more applied case, there is no way for one culture to (correctly/accurately) tell another that they are wrong about something, because regardless of the belief, they would be correct and the other culture would be correct too.

The second class of Relativism is Global Individual Relativism. This says that all facts are relative to and determined by what the individual believes. And I do mean all of them. So, for example, if this were true and a little kid believed, with all their little heart, that unicorns existed, then, for them, unicorns exist. There would be no way for another person to tell them that they were mistaken because the fact about whether unicorns exist is determined by what each individual believes. This is the ultimate, extreme, form of the phrase 'you do you, bro'.

Those last two are a little too extreme for anyone to take seriously any more. Global Individual Relativism was popular during Socrates' time, but it quickly fell out of favor because of how, frankly, crazy it is. The next version is a lot more popular now a days and it's easy to find examples of it which people could find appealing. This is Limited Cultural Relativism. 'Limited' means that it doesn't apply to all facts, rather just a small, limited variety of them. Normally, when discussing the stance, 'Limited' is replaced with the area/class of facts which are claimed to be relative. For example, one could be a Aesthetic Cultural Relativist; this means that one thinks that facts about beauty/art are determined by the culture's beliefs in those things. But, this should not be confused with 'beauty is in the eye of the beholder' because that would be Aesthetic Individual Relativism. 

The fourth and final classification of relativism is similar to the previous. This is Limited Individual Relativism. Like before, it doesn't say that all facts are relative, rather it says that only a small section of them are. As before also, 'limited' is replaced with the area which it concerns. The examples for this could be reasonable or just plain silly.  To use an example from before, Aesthetic Individual Relativism is the stance that there's no fact of the matter regarding art or beauty, rather whether or not something is beautiful is determined by whether or not the individual finds it pretty.
\begin{center}
\begin{tikzpicture}[modal]
		\node[world] (w1) [label=left:\tiny{At least some truths are relative}] {\tiny{Relativism}};
		\node[world] (w2) [label= above: \tiny{All truths are relative}, above=of w1]{\tiny{Global}};
		\node[world] (w3) [label=below: \tiny{Some truths are relative}, below=1cm of w1]{\tiny{Limited}};
		\node[world] (w4) [label=below: \tiny{All truths are relative to your culture},right=1cm of w2]{\tiny{Cultural}};
		\node[world] (w5) [label=below: \tiny{all truths are relative to you},left=1cm of w2]{\tiny{Individual}};
		\node[world] (w6) [label=above: \tiny{Some truths are relative to your culture}, left=1cm of w3]{\tiny{Cultural}};
		\node[world] (w7) [label=above: \tiny{Some truths are relative to you},right=1cm of w3]{\tiny{Individual}};

		\path[->] (w1) edge (w2);
		\path[->] (w1) edge (w3);
		\path[->] (w2) edge (w4);
		\path[->] (w2) edge (w5);
		\path[->] (w3) edge (w6);
		\path[->] (w3) edge (w7);

	\end{tikzpicture}
\end{center}


\input{culturalrelativismrachels}

\chapter{Part 2: Moral Relativism}

Moral/Ethical Relativism is a kind of limited relativism. It does not say that all facts are relative, rather it says that (only) facts concerning morality are. This stance might concern questions about whether a person is good or bad (it might say that the standards for being a good person are relative) but that stance is not popular and not the ordinary view. Moral Relativism is a very common stance that the 'woke' people might find appealing. It says that there's no absolute, objective facts about morality, what makes actions right or wrong, rather it says that the standards for rectitude are relative. They are determined by (depending on the version held) either the beliefs of the person doing the action or the beliefs of that doer's culture. Moral Individual Relativism is a rare stance these days, you are likely not going to find it on the street, so to speak, but you will likely come across  Moral Cultural Relativism (often just called 'relativism'). We will focus our attention on that latter stance, but most, if not all, of the problems regarding Moral Cultural Relativism applies to the former too. MCR is not without support, for example take these two examples:

\factoidbox{\noindent  \fontsize{20pt}{0pt}\textbf{King Darius}

Darius, a king of ancient Persia, was intrigued by the variety of cultures he encountered in his travels. The Callatians (a tribe in India) customarily ate the bodies of their dead fathers. The Greeks practiced cremation. Darius thought that an understanding of the world must include an appreciation of such differences. One day, he brought some Greeks who happened to be present and asked them what they would take to eat the bodies of their dead fathers. They were shocked and replied that no amount of money could persuade them to do such a thing. Then Darius called in some Callatians, and while the Greeks listened asked them what they would take to burn their dead fathers' bodies. The Callatians were horrified and told Darius not even to mention such a dreadful thing.\autocite{Herodotus1}}

Ask yourself, would you be willing to eat the body of your dead parent? Many of us, today, would be horrified at this notion, just as the Greeks were. But, it should also be noted that the Callatians had the same reaction to, from our perspective, perfectly normal practice of cremation.  This might make you think that morality and ethics is nothing more than norms of behavior which a society agrees to. To drive this idea home, consider this example: 

\factoidbox{\noindent  \fontsize{20pt}{0pt}\textbf{The Dip}

In some Inuit cultures, there is a common practice know as ‘the dip’ (my translation). When a child is born in the dead of winter, it is common for the mother to carve out a hole in the ice and place the child in the water. This kills the child instantly (infanticide). This is done purely at the parents’ discretion and there is no negative stigma about it. Old people also, when they became too feeble to contribute to the family, were left out in the snow to die.}

Just as before, this is, seemingly, a perfectly normal practice for those inuit cultures but from our cultural perspective, this is horrifying. In many cultures, caring for the elderly and those to young to fend for themselves is seen as a very important moral duty. Something must be very different if this culture does not have the same regards. 

These practices, which I have given, are radically different than the practices which most cultures have and, as I have hinted, might lead one to think that ethics is just a set of cultural norms, rules of behavior, with no one system better than any other. One might think that if there were a fixed set of moral rules, there would be no disagreements about morality. But, there are disagreements, so there must be no absolute moral rules. 

That line of reasoning is an argument. An argument is different from a debate in that a debate requires disagreement and the drive to prove the other side wrong. An argument is just a line of reasoning which is intended to prove a conclusion. Some arguments are faulty, they fail to adequately prove their conclusion. To see how and why, for this case, let's break it down into the most relevant parts, the premises and the conclusion:

\section{The Cultural Differences Argument and Its Problems}

Not moments ago, I gave the `ordinary language' version of the Cultural Differences Argument. But, ordinary language is full of flowers and distracting prose. To avoid those and to get to what actually matters, we extract only the supporting sentences (the premises) and the supported sentence (the conclusion). For the Cultural Differences Argument, it looks like this: 

\begin{earg}
\item[1] Different Cultures have different moral codes.
\item[2] Therefore, there's no fact of the matter regarding morality (it's culturally relative).
\end{earg}

To start us off, do you think that this is a good argument? I am going to be frank, this argument is not good. It may be persuasive, but it does not get at reality. For an argument to really prove that some fact is really the case (the 'therefore' part in this case), there is a pretty high standard. First, the argument needs to be valid, that is, if we assume that the premises (the support is true), then the conclusion needs to follow necessarily from them. For this argument, the support (the premises) concern what people believe and the conclusion concerns reality. Take this example, which has the exact same form as the argument above:
\begin{earg}
\item[1] Different cultures have different views about the shape of the Earth.
\item[2] Therefore, there's no fact of the matter about the shape of the Earth (it's culturally relative).
\end{earg}
If you accept the Cultural Differences Argument, then you will need to also claim that there's no fact of the matter about anything where there's the slightest disagreement. But, this is not the only problem for the Cultural Differences Argument and Moral Cultural Relativism, there are more:

\subsection{Cultural Oppression}

Though the argument may not be any good, all that shows is that the argument isn't good. It does not show that morality is objective. But, if we look at the last example I gave, involving the Kaelons, we see something interesting. If morality is relative to a culture, there's nothing stopping one culture, morally speaking, from oppressing people in their culture and oppressing people in other cultures. For example, if there are no objective moral facts, then there would be nothing wrong with the Kaelons declaring war on another culture for allowing someone to receive asylum. Similarly, there would be nothing wrong with the Kaelons forcing members of their culture to unwillingly commit suicide. This sort of case can be found in history as well, with the sear number of examples being too numerous to list. This is a problem which will appear again later in this class, because it's more obvious inside another argument.  

\section{The Cultural Imperialism Argument and Its Problems}

There's one other argument which one could come across in favor of Moral Cultural Relativism, this is the Cultural Imperialism Argument. To start us off, Cultural Imperialism is the view that your culture as all of the right answers and that all of the other cultures are mistaken. In virtue of this, you are justified in forcing other cultures to change to your own. Take, for example the treatment of Native Americans in America. The treatment of the Indians in India by the British. The treatment of the Latin Americans by the Spanish, and so on. Basically, the colonial practices of Europeans (and other groups, but the Europeans were some of the worst offenders) all trace back to this view, which leads to oppressive practices and forced cultural change. 

Looking over world history, this practice was wrong, they should not have done it. So, we seem to be able to use this as a premise and, maybe, just maybe, be able to get some kind of Moral Relativism from it.

\subsection{The Cultural Imperialism Argument}
\begin{earg}
    \item[1]Cultural Imperialism is morally wrong. (Intuition).
    \item[2]If Cultural Imperialism is morally wrong, then it is wrong to judge the values of one culture against the values of another. (Consequence of it being wrong)
    \item[3]If it is wrong to judge the values of one culture against the values of another, then no person is ever justified in criticizing the moral norms of another culture. (Consequence of the consequence)
    \item[4]If no person is ever justified in criticizing the moral norms of another culture, there are no non-relative moral truths (they are all relative to culture). (Consequence)
    \item[5]Therefore, there are no non-relative moral truths (they are all relative to culture).
\end{earg}

This argument is a wee-bit stronger than the Cultural  Differences argument, but it too has its fair share of problems. Can you see the big one? I'll give you a hint, look at the first and the last lines (the intuition and the conclusion). The first line gives us a non-relative moral truth, that this practice is wrong, but the conclusion claims that there are none of those. The first line disproves the conclusion and the conclusion disproves the first line. They are contradictory, the argument is hypocritical, so to speak. There are ways of strengthening this argument, limiting the scope of the Relativism, yet again. For example, one could divide the practices into internal (those concerning members of the same culture) and external (those concerning members of different cultures). If you go this route, then you could, maybe, claim that cultural imperialism is wrong because it's an external practice and only internal practices are relative. This would be Cultural Relativism About Internal Norms (CRAIN). However, this too has its issues, not concerning the arguments for it, but concerning the facts of the world. Take this example, from Star Trek:


\factoidbox{\noindent  \fontsize{20pt}{0pt}\textbf{Half a Life}

For most of the world, the elderly are treated with respect and cared for until their natural death. The Kaelon people are different. In this society, the prevailing view is that it's the duty of the 'elderly' to leave their tasks to the next generation, that forcing the next generation to care for the elderly is cruel,  and that having death come for a person  seemingly randomly is heartless. So, when a Kaelon turns 60 years old, they undergo the Resolution. In it, a great party is thrown, celebrating their life and accomplishments, and afterwards, the Kaelon commits suicide. Living past this point is seen as greedy, their time is up and their accomplishments after this point are seen as stolen from the next generation. This is expected of all Kaelons, refusing to kill yourself at the allotted time will cause even family members to be ashamed of you. If a Kaelon seeks asylum to avoid the Resolution, the Kaelons will declare war on the other culture in order to kill the person.\autocite{HalfaLife}}

We can say that the Kaelons declaring war on another culture over this is wrong, but the Kaelons forcing members of their culture to commit suicide is an internal practice, it only concerns those which are members of the same culture, so, if you think that this practice is wrong and should be stopped, we have an example of when cultural imperialism might actually be the right thing to do.  And, in fact, if we look into history, we find some cases where cultural imperialism was, in fact, the right thing to do (as in, forcing a change in a culture). For example, any culture which sided with the Allies during WW2 in order to stop the Nazis from killing the Jewish people was engaged in this sort of imperialism.  Figuring out when it's right to force a culture to change its practices is a bit tricky, but we can say that there are some cases where it needs to be done. 

\chapter{Part 3: Problems for Moral Relativism}

Even with the arguments for Moral Cultural Relativism not being good, there's still a chance, no matter how slim, that it's the correct view of the world. It could be possible that people smarter than me just haven't figured out a good argument for it. So, to really disprove moral relativism, we need to look at the consequences and see whether they are the sort of things which actually line up with realty. For this, we will look at 3 different consequences of taking Relativism seriously. If you are willing to accept all of these, then Relativism might be for you.

\section{The Criticism Problem}

\begin{center}
\textbf{We could no longer say that the customs of other societies are morally inferior to our own}
\end{center}

This is the point that cultural relativists really want to stress and it is one that people who like this avenue believe. But here are some of the things that follow:
\begin{earg}
    \item[]We cannot criticize cultures with slavery (pre-Civil War South).
    \item[]We cannot criticize cultures with anti-Semitic views (the Nazis).
    \item[]We cannot criticize cultures with rigid caste systems (found the world over, throughout history). 
\end{earg}
If we took cultural relativism seriously, we would have to regard these social practices as also immune from criticism. Any country entering WW2 would have done wrong (so long as they did so to stop the practices). This issue was brought up with the Kaelons example before and does warrant retelling.\footnote{There are several other examples which can be found where forcing a change in a culture seems like the right thing to do. If you take an anthropology class, you may encounter examples involving female circumcision or, by another name, female genital mutilation. Forcing them to abandon this practice seems like the right thing to do. Forcing the change, though the right thing to do, will often require that it's done in the right way. It should be done gradually, recognizing the reasons why the culture does the practice and then helping to eliminate those reasons.}

\subsection{The Argument for The Criticism Problem}

\begin{earg}
    \item[1] If MCR is true, then morality is relative to culture. (This is basically the definition)
    \item[2] If morality is relative to the culture, then there is no standard to gauge the morality of a culture.
    \item[3] If there is no such standard, there is no way to compare one culture to another (morally speaking)
    \item[4] If there is no way to compare them, then there is no way to say that the Nazis were bad.
    \item[5] Therefore, if MCR is true, then there is no way to say that the Nazis were bad.
\end{earg}

So, Moral Relativism implies that one culture could never, accurately, pass judgement on the practices of another, for good or for ill.

\section{The Sheeple Problem}

\begin{center}
\textbf{We could decide whether actions are right or wrong just by consulting the standards of our society.}
\end{center}

Cultural Relativism suggests a simple test for determining what is right and what is wrong: All one need do is ask whether the action is in accordance with the code of one's society. To help picture this, imagine that whenever the standards for being a culture are met (that's another problem all together, "what does it take for a group to be a culture?"), a big book falls from the heavens with all of the norms of that culture written down. So, if you ever have a question about whether an action is right or wrong, you would just need to open the book, find the relevant rule and follow it like a sheep.  For example:  Suppose that a person in the deep south (Pre-Civil War) was curious about whether slavery was permissible. All they would need to do is ask whether it fit with the code of the society. If it did, then they would be OK with having slaves. People often think that believing in Moral Relativism is in some way liberating when in fact, it's imprisoning.  Not only can’t we criticize other cultural codes, but we can’t criticize our own. (All human rights movements would be incorrect according to this).

\subsection{The Argument for The Sheeple Problem}
\begin{earg}
    \item[1] If MCR is true, then morality is relative to culture. (This is basically the definition)
    \item[2] If morality is relative to the culture, then whether an action is right or wrong is determined by your cultural norms.
    \item[3] If those are determined by the norms, then all you need to do is consult the standards of your culture to determine morality.
    \item[4] So, if MCR is true, then all you need to do is consult the standards to determine morality. (Intermediary conclusion)
    \item[5] If morality is determined by the standards, then acting against those standards is morally wrong.
    \item[6] All Human/Civil Rights movements act against the standards of the culture.
    \item[7] Therefore, If MCR is true, then all Human/Civil Rights movements were/are morally wrong.
\end{earg}
So, Moral Relativism implies that you need to be a sheep to your culture, that you shouldn't object to a practice and blindly follow it, no matter how bad.\footnote{To clarify this, there is a general norm which says that it's wrong to criticize a person for doing something morally permissible. In seeking to change a norm of a culture, the person is criticizing them for doing something which, by their culture, is permissible. This makes seeking to change the norm wrong. In a more applied case, the protests against segregation in the US, according to Moral Relativism, were wrong because they violated the norm, segregation, which they were trying to change. There has never, and likely could never, be a case where trying to change a norm in a culture does not violate a norm in that culture.}

\section{The Progress Problem}

\begin{center}
\textbf{The idea of moral progress is called into doubt}
\end{center}

We often claim that certain societal changes are for the better and of course, we often claim that some of them are for the worse. Take, for example, the history of Women's Rights. This is such a universal problem that I really don't need to point to any individual one. Throughout most of Western history the place of women in society was narrowly circumscribed. They could not own property; they could not vote or hold political office; and generally they were under the almost absolute control of their husbands. Recently much of this has changed, and most people think of it as progress. Progress is where the changes move something closer to an objective standard (in this context). We say that we are making progress towards fixing a car, finishing an essay, or any other project with a fixed end-point. If Moral Relativism, of any sort, is correct, then there's no objective standard which we are moving towards, so there's no way of saying (correctly) that we have made progress. Similarly, different people at different times are, for all intents and purposes, different cultures (so long as standards have shifted). So, like with the first consequence, we can't say (correctly) that the standards had at one time are better/worse than the standards had at a different time. So, in the same way, if MCR is correct, then there would be no progress, at all.\footnote{For another way to think about this, we judge progress by comparing how the situation is and how it was according to some end goal. If Moral Relativism is correct, then there is no end goal for the comparison and there is no means of comparison even if we devise some other scale. For example, take two pieces of paper and cut one of them to be a different length than it was. Putting them side-by-side, you could say that one is shorter than the other. But could you say that one is shorter than the other if you were only ever able to see one of them at a time (and you didn't cut them)? Could you do so if you were incapable of remembering the other piece? You couldn't. Moral Relativism takes away all abilities to compare the moral standings of cultures, not only different cultures at the same time but the same culture at different times.}

\subsection{The Argument for The Progress Problem}
\begin{earg}
    \item[1] If MCR is true, then morality is relative to culture. (This is basically the definition)
    \item[2] If morality is relative to the culture, then there is no standard to the morality of a culture.
    \item[3] If there is no such standard, there is no way to compare one culture to another (morally speaking)
    \item[4] Different people at different times, if the norms changed, are, for all intents and purposes, different cultures.
    \item[5] So (from 1, 2, 3, and 4), if MCR is true, then there is no way to compare people from one time to another (morally).
    \item[6] If there is no way to compare people from different times, there can be no way for us to say that we are better than we were (made progress).
    \item[7] Therefore, if MCR is true, then there is no way to say that we have made progress.
\end{earg}

\chapter{Part 4: Other Problems for Moral Relativism}

\section{Why There is Less Disagreement Than There Seems}

The Cultural Differences Argument centers around the idea that the differences between cultures are insurmountable. It holds that the differences are fundamental differences in values not just differences in perspectives or views about the nature of the world. This, however, is a rather strong claim and can be easily debunked once we do a little digging. Many robust claims like this are the product of what I will call intellectual laziness. This is where the person holds the stance because they are too lazy to learn something more about the subject and then that laziness lets it fester into an unquestioned belief. So, let's use a simple example to illustrate this. Suppose that news broke that some far off culture holds that birds are never to be eaten. Eating or killing a bird is equivalent to killing or eating a human person. On the face of it, without doing much further digging, one could think that this amounts to a fundamental difference in how they place values on the world. But, it would be intellectually lazy to let it rest there. If we do a little digging, we find that the culture actually believes in reincarnation and that good people are reincarnated as birds so that they can hear prayers and fly back and forth between Earth and Heaven. This changes the perspective we should have about the culture. It's not that they have a fundamentally different value system, rather they disagree with us about the role of birds and the nature of the soul. For them, killing a bird is the same as killing a good person and we disagree about whether a bird was once a good person. 

\section{There are Norms Shared by All Cultures}

One mark against Moral Relativism, aside from the lack of real disagreement about values, is that there are certain norms which are shared by all cultures. These norms could come from various different sources, but typically, for a culture to survive through time, they are going to need to have these. There is sort of a natural selection process which weeds out cultures which lack or discredit these features. The first of these features is that any culture is going to need to place a high value on raising or caring for their young. Even with the Inuit cultures, which practice The Dip, if an infant survives, then it is very well cared for. This is because a culture which does not care for its children will likely not survive past a couple of generations or those who adhere to that custom will not survive past a few generations and the culture will change to one which cares for the young. Failing to care for the young or infanticide will need to be the exception rather than the rule. Next, all societies will need to place some value on honesty. Lying, cheating, and deceit, if allowed to propagate in a culture will quickly be the undoing of that culture and those who are honest (in the right times and conditions) will survive to continue a new culture which values the honesty. And, for a third example, all cultures will need to have some sort of ban on needlessly killing other members of that culture. This should seem obvious but any culture which allows for murder and killing to go unchecked will likely not survive through time and will eventually be replaced by one which has such a ban on murder. 

\section{But What is a Culture?}

So far, I have danced around the notion of being a culture. I use the term and expect a certain intuitive understanding. But, there are questions which we should ask. Moral Relativism bases its claims about morality on the notion of a culture, so we should ask what it takes to be a culture. One could go a few different routes about what it takes for some group of people to be a culture and we could even ask about whether there even are cultures. The first way we could define a culture is to have it be a collection, or set, of norms, or rules of behavior which are followed by a certain group of people. We have further worries about the number of people and so on, but this is a reasonable place to start. Defining a culture as a set of norms, however, runs into a bunch of issues with vagueness.

Suppose that there is a group of people, say 1000, who all believe in a certain collection of norms. We will say, just assume, that this is a culture, C. Now suppose that as time progresses some people disagree with one of the norms in C, for example, about whether you should pour the milk or the tea first. This movement grows and now 500 believe that the milk should go first and 500 believe the tea should go first. Do we say that C has split into two different cultures or do we say that the rule isn’t settled? If we say that C has split over one rule, then any disagreement would mean that the Cs would split again, and again, and again, eventually leading to 1000 different cultures. Every person would be a culture of one and, morally, one could never criticize or correct another because they are a different culture.  In real life, this is far more messy and even more disastrous for Moral Relativism.  If we say that the rule in the culture is undecided, we get a similar result. Any disagreement within a culture would result in that culture having the rule be undecided. So, eventually, there would be no norms of the culture at all which would destroy the culture from ever existing. As a result, if we say that disagreement in a culture is allowed, then there are no cultures at all and if we say that any disagreement splits the culture, then all people would be a culture of one. 

That example is an issue inside of the culture, but if we expand it and use it to compare cultures, as is common, then we run into similar worries. Suppose that C meets another culture D. D has many of the same norms as C except one, which is different. If we say that, despite appearances, they are actually the same culture, C’, we get an interesting snowball. C’ shares many of the same norms as another culture, E, so they are the same culture. This slowly builds until there’s a point where there aren’t different cultures, but rather there is only one culture, the human culture, with minor regional variations.

One could, attempting to escape this, define a culture merely as a certain group of people when they hit a certain number. The notion here is that once you have a group of a relevant size, then the norms will organically emerge. There are issues with this such as geographic distribution and stuff like that, but we can put those aside.  We could say that a culture is a group of so many people. But, is there a hard line for the number of people it takes to be a culture? What if a group has exactly enough people to be a culture and one or two die? Did the culture die too? Morality should not be so fragile.

\chapter{Part 5: If Moral Relativism is wrong, What's Next?}

If you are willing to say that moral cultural relativism is wrong, and, given the general responses to the various problems with moral relativism, you likely are willing to say that it's wrong, we have some problems. Namely, what is moral? Who decides who's correct in a moral conflict? Where do moral truths come from? Here I will cover some of the basic questions which former relativists tend to give and the general responses (sometimes, the response will need to be more particular).
\section{If culture's don't choose, what is morality?}

There are a few ways to go about replying to this question. Morality/ethics centers around 'should' questions of a certain kind. When we ask questions like these, for example "what should I do?", "should you call a doctor?", "should you get the assignment in on time?",  there are two different senses, which may or may not overlap, depending on context and relevance. Something is moral/ethical when it's the correct answer to the question "what should I do?", when we aren't talking about practical cases (a practical case is one where you are asking about the correct way to perform some task, like changing the oil on your car). The cultural moral relativist would claim that the correct answer to the question in non-practical cases (for example, should I flip the switch in the trolley problem?)\footnote{You will see me use the Trolley Problem as an example in a few different places. It essentially boils down to asking whether it is permissible to kill one person in order to save 5. There are many versions of it out there and some might make you think that it's wrong while others will make you think it's the right thing to do. That's why it's a problem, after all.} will depend on your culture and the norms associated with it. The moral objectivist, a moral realist who is also a non-relativist, would claim that there's an actual answer to this question, which is not determined by your culture. The next module of this class covers some of the ways which philosophers have answered how to answer that question, without being relativist. Ethics is, at it's core, trying to find the correct answer to this question.

You will notice that when I give examples of cultural norms, most of the time, these are cases where the culture got the answer wrong. Ethical theories are basically hypotheses about the correct way to get the answer to the question. Some theories point to absolute, objective duties which a person has (typically, it's wrong to be irrational; acting in a certain way is irrational, therefore those ways of acting are wrong). Others point to the well-being of those affected (if an action makes the affected better off than otherwise, that's the right action). Others still point to some exemplar of morality, some person who has the perfect character and then asks "what would that person do?". You may have heard that kind of thinking before with 'WWJD'.  
\section{Who decides who's right in moral conflict?}

A moral conflict is a case where two or more people/groups disagree about the answer to the "What should I do?" question, in the relevant sense. The moral cultural relativist has a simple way to answer it, almost too simple, "if it's two groups in the same culture, the cultural norms settles the conflict, if it's two cultures, there's no real conflict." But this just does not seem right, if you recall the last page, were this true, all civil rights movements would wrong, because the cultural norms would settle the conflict in favor of the oppressing group. Very few of us would want this. So, that's out, but is there a non-cultural way to get the answer? This is the quest for the moral realist. It's not going to be based on belief (as the relativist points to).

Recall the Cultural Differences Argument. This argument relies on moral conflicts/disagreements. But, do we always settle debates like that? There may be disagreement about something, anything (for that matter), but it doesn't follow that there's no fact of the matter. If there's a conflict about math, we don't say that there's no answer, we consult the rules of mathematics (as they were discovered, not invented). When there's a conflict in science, we don't say that there's no answer, we perform experiments and discover the truth. For the moral realist, settling moral conflicts is more like settling a conflict in math or science than one in art. Realists perform experiments and consult the rules of morality to settle the debates (and the experiments either further support the rules or give examples of amendments which should be made). The quest to get the non-relativist rule to settle moral conflicts is hard, but it does not seem impossible, we do make progress in it. There's even a scientific-style method for figuring out which ethical theories should be applied and/or how they should be amended (as we will see in the next module).

In short, no one decides who's right in a moral conflict, it's just a fact that one is right and the other is wrong.
\section{Where do moral truths come from?}

Moral truths are the answers to the "what should I do?" question (in the relevant sense). The moral cultural relativist has a simple, again almost too simple, answer "from the beliefs of the culture." Yet again, how can we possibly say that a culture has the wrong answer to the question when the culture chooses the answer? Like the issues before, this just can't be right. So, the question for the realist is just that, where do they come from? 

For the realist, this is like asking "where do scientific truths come from?" or "where do mathematical truths come from?". Some moral objectivists point to God and say that all truths flow through Him, moral truths included, saying that God made them (an easy way to get all-knowing and all-good).\footnote{We will be encountering this theory, called Divine Command Theory, later in the course. But, for a primer, that theory claims that an action is right if and only if God commands it and wrong if and only if God forbids it. There are many problems with this theory, which we will see.} Others point to the sort of creatures we are and our place in nature, these give us moral truths. Others still point to abstract notions of well-being or rationality.  And even more say that moral truths are things which just could not have been otherwise, they just always are, they didn't "come from" anything. There is some debate about where they come from.

But, don't lose heart! There's just as much debate about where mathematical and scientific truths come from. For example, in the case of math, we could claim that the truths are necessary, they could not have been otherwise. We could claim, with just as much evidence, that the truths are constructs of the definitions of the terms which we are using overlapping in consistent ways (1+2=3 because of the definitions of the terms and operations). And we could also claim that the truths came from God.  As it turns out, in philosophy, the most common stance (for where mathematical truths come from and where moral truths come from) is that they always were and always will be.
\section{What about forcing them onto another culture?}

This is a hard problem, because one of the core intuitions behind the Moral Relativist, and one of the big reasons why the stance is contradictory, is that it’s wrong to force your morals onto another. The Moral Objectivist will claim that under the right circumstances, it’s morally required that you force another to change. To some, this feels wrong. This feeling, often, gets traced back to the horrors of history, which I used as examples for the intuition behind the Cultural Imperialism Argument, where one group imposed themselves on another. At the same point, we need to look at history as a whole. There have been times when it was actually correct for one group to come in and force a change. In those cases, the imposing groups had real morality on their side and in the cases where it was wrong, we can say that they were mistaken about morality. For example, the Nazis in WW2 were just wrong about morality and the groups which came in were right. So, the Moral Objectivist just needs to be sure that their moral theory is getting the right answer in the case. This is an epistemological worry for the objectivist, not an ethical, meta-ethical, or metaphysical one. 

Interestingly, when we look at our own history, we find things which we are ashamed of, things which we condemn the previous generation for. How and when this is correct is easy for us to see. Since different people at different times are, basically, different cultures, the exact same, or a similar, thought process applies when condemning or appreciating the behaviors of another culture. Kwame Appiah in What Will Future Generations Condemn Us For? gives us some compelling tests for when future generations will be ashamed of our practices. Appiah gives us three tests. These tests don't tell us what we should do, rather they tell us that something is wrong in the practice we have and it needs to change. We can use the hindsight of the future generations, which is 20-20, to tell us what we need to do today.\autocite{Appiah1} 

The first sign that future generations will be ashamed of our behavior is that we already know the arguments against the practice. Most of the time, when there is a morally questionable practice, the arguments against it are actually quite old. Very, very, rarely is there a moment of moral clarity which changes the world. For example, take the consumption of meat or the meat production industry. The arguments against the various practices in farming meat are centuries old and this is a sign that something is off morally and future generations will feel shame for the practices we engage in. For another example, arguments against police brutality and the militarization of the police force are centuries old. In fact, many of those arguments are why most officers in England do not carry firearms. This is a sign that something needs to be done about police brutality. 

The second sign is that defenders of the practice use non-moral reasons to justify or defend the practice. The defenders point to religion or tradition, they point to 'human nature', or they point to some sort of economic or cultural necessity. Moral reasons should always trump non-moral reasons. So, the defenders are, in a sense, floundering. In the case of tradition, the defenders will say things like "it's always been this way, why should we change it?" Future generations will likely read about these arguments in textbooks and feel shame because of how nearsighted their ancestors were and proud of those who fought to change it. In the case of  'human nature', very rarely is that a good mark for morality and we should strive to better ourselves rather than be fatalistic about out lot in life. And with cultural or economic necessity, they make claims like "the economy couldn't survive such a change!" Future generations will likely express that any system which only survives on injustice has no right to survive. 

The third and final test is that of strategic ignorance. Most of the time, also, when people are accustomed to a system and like it, they are resistant to changing it. Not only will they resist changing it, but they will intentionally or subconsciously ignore some of the factors which point to the practice being bad. These factors will become harder and harder to ignore for future generations and eventually they will change the system and feel shame for our inaction. For example, economic inequality is strategically ignored by many or they emphasize the rags-to-riches stories which support the system. They ignore the fact that the odds of pulling one up by their boot straps is decreasing with every generation and they ignore the fact that the opportunities to pull oneself up are not evenly distributed. This is a sign that there is something in this system which future generations will be ashamed of.