\part{Moral Objectivism (Non-Relativism)}
\label{ch.mod2}
\addtocontents{toc}{\protect\mbox{}\protect\hrulefill\par}
\chapter{Part 6: Objectivism and Absolutism}
Since there are some very serious problems with Relativism, philosophers and others have moved to a different way of thinking about Ethics. They asked themselves "what if ethics is absolute? What if there are certain things which are wrong regardless of the culture?" This gives us \gls{Moral Objectivism} (not to be confused with Ayn Rand's Objectivism, which should just be called Stuff-ism, we will cover this when we cover Ethical Egoism). Moral Objectivism is a meta-ethical stance, like Moral Relativism. It says that there are absolute, objective, moral truths. These truths apply to everything, like the truths in mathematics and the other sciences.  I personally prefer the term ‘objectivism’ rather than ‘absolutism’ for this subject (all of them actually), but there is a slight difference. Like the other stances, it can be divided into global and limited. However, it does not need to be further broken down from there. For this class, we are focusing on moral objectivism. This is the stance that there are absolute, objective, rules about morality. They aren’t relative and apply to ALL people. When we make a claim like ‘lying is wrong’, we are saying something which is either true or false and whether it is true or false is not relative.

\newglossaryentry{Objectivism}
{
  name=Objectivism,
  description={The stance that there are absolute, objective facts about a given topic.}
}

\newglossaryentry{Moral Objectivism}
{
  name=Moral Objectivism,
  description={The stance that there are some absolute, objective facts about morality, right and wrong.}
}


Moral Objectivism could be divided into two different, closely related, stances. One could be a \gls{Moral Absolutism}. This is the stance that the moral truths, the rules, are objective but it goes one step further and claims that the moral truths are without exceptions. They are absolute. There are many moral rules which we apply in our daily lives. One should always obey the law, one should not lie, one should not steal, one should not cheat, and one should not harm others. If moral absolutism is correct, then these rules (if they are the real moral rules) are without exceptions. There is never a case where lying is the morally right thing to do, for example. For example, if the claim “killing is wrong” is true, then it is true regardless of context. So, killing one person to save 5 is wrong. 

\newglossaryentry{Moral Absolutism}
{
  name=Moral Absolutist,
  description={The stance that there are absolute, objective moral rules which are exceptionless and absolute. For any, if they hold in one case, they hold in all cases}
}


Moral Situationalism is different. This holds that moral truths are objective, but they apply differently according to context/situation. \gls{Moral Situationalism} is happy to say that most of the time the moral rules we apply in our daily lives are accurate, but they are not absolute or without exceptions. If we take the previous list for examples, Moral Situationalism is willing to say that most of the time one should obey the laws of the land but there are or could be cases where the laws are unjust and should not be followed. If obeying the law were an absolute moral rule, then it would be impossible for a law to be unjust. So either moral absolutism is not correct or obeying the law is not an absolute moral rule. Typically, a situationalist theory has at least one absolute moral rule involved, but this is typically very abstract and not all that useful in a practical situation. It’s used to determine which rules apply in a context. So, for example, most of the time, stealing is wrong, but there could be cases where it’s the right thing to do because the rule doesn’t apply. For another example, lying to your spouse about what you did over the weekend is wrong but lying to an axe-murderer about the whereabouts of your friend is right. Situationalism can be seen in other fields aside from Ethics. If a little kid points to a silver car and says that it’s silver, they would right. But if they point to a different red car and say that it’s silver, they would be wrong. The situation, the car in this case, is different so whether the kid is right could be different. Theories about what makes the situation different are numerous and some, but not all, are quite intuitive, but this is not a class about Philosophy of Language.

\newglossaryentry{Moral Situationalism}
{
  name=Moral Situationalism,
  description={The stance that there are objective moral rules but they apply differently in different situations. There is typically an absolute rule which determines whether the more practical ones apply}
}


There are several reasons to be a moral objectivist aside from getting out of the problems with Relativism and still being able to claim that there are certain things which you shouldn't do. But those problems are negative arguments. They say "well Relativism is out, so you must go with Objectivism." This doesn't really convince most people, however, there are some tests to tell whether you actually, in your heart of hearts, think that a subject is relative. These are our more positive arguments, our evidence that people really think morals are objective.

\input{objectivismEnoch}
\chapter{Part 7: The Tests for Objectivity}
\section{The Spinach Test For Objectivity}

This is a pretty straight forward test for whether you actually are a moral relativist. But, as a fair warning, philosophy kills jokes. As an example, and where the test gets its name, lets look at the following joke:

    \factoidbox{A kid hates spinach, later, he says that he’s glad he hates spinach. When asked why, he says “because if I liked it, I would eat it; and it’s yucky!”}

More often than not, we will take this as funny. And, in a lecture format, if I present it right, I will get a laugh. But, the test is not whether or not you laugh at this joke, but rather whether you laugh at jokes of this form/structure, with different things put in the place of 'spinach'. Whether you laugh is the answer to whether you think the topic is relative. You can take the joke and put it into this general form:

\factoidbox{A person believes X. Later, they say that they are glad/happy that they believe X. When asked why, they say "if I believed otherwise then I would Y, but that's just wrong/yucky/boring.}

For example, let's try it with entertainment and other foods:
\factoidbox{
    A kid hates watching golf. He says that he’s glad he hates watching golf because “if I liked it, I would watch it, and it’s boring!”

    A metal-head says that he’s glad he hates country. When asked why, he replies ‘because if I liked country, I would listen to country, and that’s just bad music.’

    George H.W. Bush hates broccoli and says that he’s glad he hates broccoli. The reason? “Because if I liked broccoli, I would eat it, and it’s gross.”}

These are the same, if presented right, they will get a laugh. But, will this joke work for anything which I plug into the joke's formula? What if I plug in something which is more absolute? Take these examples:
\factoidbox{
    A 20th century man believes that the Earth is round. He claims that he’s glad that he’s a 20th century man because “If I grew up in the first century, I would have believed that the Earth is flat, and that's just wrong.”

    A 20th century man believes that women should have the right to vote. He claims that he’s glad that he’s a 20th century man because “If I grew up in the 19th century, I would have believed that women shouldn’t vote, and this is wrong.”

    An Iraqi woman who grew up in the 2000s believes that women are not equal to men. She claims that she’s glad about this because “if I grew up in the 60s and 70s, I would have believed that we were equal, and this is wrong.”}

There, people tend not to laugh, the joke doesn't work, so the topic in question (whether the earth is the center of the solar system) must be objective. Usually, our stances about liking or disliking something are totally up to us. If we dislike something, the reason we shouldn’t have it is exactly because we don’t like it. So if we're imagining a hypothetical scenario in which you do like it, then you no longer have any reason not to eat it. This is what the child in the first example gets wrong: He's holding fixed his dislike for spinach, even in thinking about the hypothetical case in which he likes spinach. But because these issues are all about him and what he likes and dislikes, this makes no sense. But physics is different: What we want, believe or do – none of this affects the earth’s orbit. The fact that the earth revolves around the sun is just not about us at all. So it makes sense to hold this truth fixed even when thinking about hypothetical cases in which you don't believe it. And so it makes sense to be happy that you aren’t in the Middle Ages, since you’d then be in a situation in which your beliefs abut the earth’s orbit would be false (even if you couldn’t know that it is). And because this makes sense, the joke isn't funny.

The Spinach Test is where we take the joke and change the thing the person believes. If it’s funny, then we have evidence that it is subjective, totally up to us. If it isn’t funny, then we have evidence that it is not subjective, but objective.

\section{The Disagreement Test For Objectivity}

Often, in real life, we get into disagreements with people. Sometimes these are serious, like whether refugees should be allowed in the country or whether global warming is occurring. Other times they are just silly, like whether broccoli is yucky or whether dark chocolate is better than milk. The second test for relativism about a subject concerns the nature of such disagreements. For this test, you need to  imagine, or actually get into, a disagreement about something. Ask yourself what the disagreement feels like, what are you using in the disagreement (are you calling on facts or preferences?). For example, take these examples:

\begin{tabular}{p{1.5in}|p{1.5in}}
Arguments A&Arguments B\\\hline
A disagreement about whether the Earth is flat.&A disagreement about whether pineapple should go on pizza.\\
\hline
A disagreement about some mathematical principle.&A disagreement about how enjoyable Buffy the Vampire Slayer was.\\
\hline
A disagreement about whether refugees should be allowed in the country.&A disagreement about the best Metal band.\\
\hline 
A disagreement about whether slavery was bad.&A disagreement about the direction of a toilet paper roll.\\
\hline
A disagreement about whether God exists.&A disagreement about whether dark chocolate is better than milk.\\
\end{tabular}

These disagreements are very different. Think about what it’s like to be in those kinds of disagreements. In the case of broccoli or chocolate, you are trying to state your own preference and then, maybe, trying to get the listener to come to their senses and change theirs. However, when we are engaged in a disagreement about, say, global warming, the point of the argument is not to change the other person’s mind or to state how you feel. Rather, it is to get at the truth. Looking above at the examples, do the ones in the left column feel the same? Do they all feel like facts? Similarly, do the ones in the right column all feel the same? Do they feel like preferences? If the feeling of the disagreements in the left column all feel the same, then you are an objectivist about morality, in your heart. You can also ask yourself what are the things you are using in the disagreement, are you calling on your preferences or are you calling on the facts about the world? If, when you are having the disagreement, you are calling upon facts and/or data, then you think that the subject is objective, it's just not up to you. If you are calling upon preferences or feelings, then you think that it's relative or subjective. 

We can be even more particular about the case, in the sense of morality. If you are calling upon facts, then you think it's objective but if you are also calling upon facts in the particular case then you think the it's situational. So, for example, if you are in court for stealing a loaf of bread and you use facts about your individual situation to mitigate or relinquish the guilt, then you are a moral situationalist, because the fact or norm 'stealing is wrong' changes according to context/situation (with some more general principle determining whether it applies). If, on the other hand, your judge/prosecution use general facts and ignore the contextual factors, then they are an absolutist about stealing. Using another example, if you are in a disagreement about refugees being allowed into the country and say "no, absolutely not ever" or say "yes, always, let them pour in", then you hold that the rules about refugees being allowed into the country are absolute. If you say something like "in this particular situation, yes" or "in this case, no" or "it depends on who we are letting in", then you hold that the rules about refugees are situational. This becomes more obvious as the disagreement drags on and the sorts of facts you call upon to defend your stance.  

So, for this test about whether you are a relativist about some subject, you need to ask yourself what it feels like to be in a disagreement about the subject. Imagine that you are having a disagreement with someone about the morality of abortion. How does that feel? Does it feel like you are having a discussion about pineapple and green-olive pizza? Or does it feel like you are having a discussion about global warming?

Yet again, very rarely do we feel like disagreements about morality are disagreements about preference. When we have those debates, it feels like they are about facts.

\section{The Counterfactual (What-if) Test for Objectivity}

Like the other two, this one is a bit different. It involves what are called counterfactuals. They are called this because they are counter to the facts. We use these sort of statements all the time without realizing it. Any time you ask yourself what would happen if something were the case, you are using a counterfactual. These are insanely useful in philosophy as well as in science. There's an entire cottage industry in philosophy dedicated to coming up with how we use, accept, and reject these kinds of statements. That being said, since we use them so much, your gut intuition will suffice. For this test, we will present the subjects as yes-no 'what if' style questions. Such as, ‘what if I drove a 120MPH on a 60MPH highway, would that be dangerous?’ It is not true that I drive that fast, but I am trying to figure out what would be the case if I did. For this test, ask your self what it would be like if the vast majority of people believe opposite to you, would you still be right/correct? Here are some examples of what-if style questions which show this test at work.

\factoidbox{
If the vast majority of people believed that politicians were actually aliens, would they still be human?

If the vast majority of people believed that global warming wasn't happening, would it still?

If the vast majority of people thought that gender discrimination was fine, would it still be wrong? 

If the vast majority of people thought that the Earth was flat, would it still be round? 

If the vast majority of people thought that Philosophy was pointless, would it still be valuable? 

If the vast majority of people thought that pineapple on pizza was delicious, would it still be gross? 

If the vast majority of people wore top hats and thought they were cool, would they still be out of fashion?

If the vast majority of people thought that Buffy was the worst show ever, would it still be great?

If the vast majority of people thought that sweet potatoes were tasty, would they still be gross?

If the vast majority of people thought that spinach was gross, would it still be tasty?}

For each of these, there should be a yes or no answer, one which you can easily give. In some versions of this, I have presented it as "all people" rather than "the vast majority of people", but that has lead to some confusion. As before, the normal responses to the questions in the left column are all 'yes', while the answers to the questions in the right are normally 'no'. This test, if the question is formulated correctly, is the most definitive, I think. It points out the fact that objective truths are not up-to-us, they will be the case regardless of whether or not people discover them. 

The steps to preform this test on yourself are pretty straight forward. First, take a stance about something which you think is true. For example, you can have that slavery is bad, that pineapple on pizza is good, that the Kaelons shouldn't force their elderly to commit suicide, that the Earth is round, or something like that. Next, you imagine a case where the vast majority of people believe the opposite. And, finally, ask yourself whether, in this strange world, you would still be correct.  If the question is phrased correctly (it's possible to phrase them so that you will get, consistently, the opposite answer) and if you get the answer 'yes', then the subject matter is objective and if you get the answer 'no', then the subject is relative. So, for example, would top hats be out of fashion if the vast majority of people wore them and thought they looked cool? No, so fashion is relative. There was a time in this country where the vast majority of people thought that indigenous peoples were less than those of other descent, were they wrong? Yes, they were, so the morality about racism is objective. If the vast majority of people thought that Philosophy was pointless, critical thinking would decline and people would end up in a great intellectual regression. Those facts alone show that the value of philosophy isn't determined by what people believe about it, people can be mistaken about the value of a subject. 

\chapter{Part 8: Why Does It Matter? (Moral Skepticism)}

So far, for the three tests for objectivity, we have that morality is objective. We don't laugh at jokes about morality (at least serious cases of it), we feel like we are getting at facts when we disagree about morality, and ultimately, morality doesn't change with the whims of the majority (perspectives do and those could be more or less accurate). But, why would we want morality to be objective? And can we know about moral truths?

On its face, we want morality's objectivity to conform to our deeply held intuitions. We want it to explain why we don't laugh when jokes are made about it (it's too serious and not up to us). We want it to explain why we point to facts when we disagree about morality (moral facts are facts, not preferences, and you don't bring feelings to a fact fight). And we want it to explain why it doesn't change with the whims of the majority. Objectivity does all of those things, morality is just not up to us, we don't determine it based on our feelings. For example, whether the Earth revolves around the sun is not determined or changed by what people believe. Moral truths are independent to how we feel about them. We seek to discover moral facts, we don't invent them.

There are some other reasons we want morality to be objective, which relate to the objections against moral relativism. First, objective morality gives us a means to compare different cultures and people and then say how one is better or worse morally than another. We can use this same comparison to determine how one should improve. We can use it to look inward and determine how our culture or ourselves need to improve morally. We are not sheep to the views of our culture and Moral Objectivism gives us something beyond our culture to guide us. And third, Moral Objectivism gives us a standard by which to judge our progress, we can use this to figure out where we need to go and what we need to do morally to be better.  
How Do We Know?

As I said, we do seek to discover moral truths, but how do we know when we get them? The scientific method, for example, is a relatively new creation for discovering truths about the world and yet people have been trying to figure out morality for far longer than that. Discovering moral truths is hard, but not impossible. Those who hold that it is impossible are called \gls{Moral Skepticism}. Skepticism, in Philosophy, is the stance that knowledge about some subject is impossible. You could be a global skeptic or a limited skeptic, much like how you could be a global or limited relativist. It is worth noting that Moral Skeptics are not making a claim about whether there are moral truths, rather they are making a claim about whether we can know them, which is different. For example, there are certain numbers in mathematics which are so large that one can't possibly know what the first digit is, but there are still facts about what number it is.

\newglossaryentry{Moral Skepticism}
{
  name=Moral Skeptics,
  description={The stance that knowing whether an action is right or wrong is impossible}
}


People become moral skeptics for several different reasons. It should be noted that all of these are bad arguments and each can be easily pushed aside. They point to disagreements, empirical science, and degrees of certainty. To start off with, here is an argument against the idea we can know anything in Ethics using disagreements:

    \factoidbox{People and societies disagree about ethics. If it was possible to know the truth about this, then really smart people would come to it eventually. They haven’t. So, it must not be possible for us to know about it.}

This should remind you of the Cultural Differences Argument. The core intuition which is being used by those who pose this objection is that if we haven't found the answer to a question (yet), there must not be an answer to it. This is flawed on many different levels. First, right now we disagree about ethics, and if we can know about it, smart people will get to it someday, and, sure, we haven’t yet but that doesn’t mean it’s impossible. It’s just a really hard problem. For most of human history, people disagreed about what the stars were, but we now know what they are, it was a hard problem and we found ways to figure out the answer. Ethics is no different, for most of human history, we have disagreed about certain moral rules. Those rules will be settled, not because we get grinded down and all agree, but rather because we discover a moral truth. 

The next argument against the possibility of moral knowledge comes from a robust, strong, faith in the empirical sciences. Philosophy itself is a science, just not empirical. Here is a basic outline of the sort of argument used:
\factoidbox{
    The empirical sciences are supposed to give us knowledge about the world. They say nothing about ethics. So it must not be possible for us to know about it.}

For this one, there are a few different replies, and this is my preferred: The empirical sciences are SUPPOSED to give us knowledge about the world and you’re right, they don’t say anything about ethics. But, there’s more than one way of getting knowledge about the world and the empirical sciences are just one of the ways. The empirical sciences can't answer questions about the nature of God, can't answer questions about beauty, can't answer questions about truth (because that's meta to their fields), and they can't answer questions about morality. Empirical science is just one way of getting truths about the world, but it isn't the only way. The history of philosophy is one of specialization. There are some questions which have objective answers but can't be answered using the methods in the empirical sciences. Philosophy, as a field, handles the questions which her children can't.

The last argument gets a little closer to home. Skepticism concerns knowledge and as such arguments for it for some subject should touch on the standards for knowledge. 

    \factoidbox{Knowledge is a really high degree of certainty. With any action we take, there’s always some uncertainty about whether it’s the right one. This means that we can’t know whether the action we are taking is the right one.}

For the third issue, there is a simple response: Alright, I agree that knowledge is a high degree of certainty, and there’s always going to be some uncertainty about the action I take. But, how much certainty is enough for knowledge? What if I am 99\% sure that this is the right route? I am still unsure, but I am very certain. We can't be absolutely certain about anything in life, morality is no different, we can only be really really sure. Going this route not only denies the possibility of knowing moral truths, but it also messes with the possibility of knowing anything else.
